Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
getwd()
pf <- read.csv('eda-course-materials\\lesson3\\pseudo_facebook.tsv', sep='\t')

qplot(x=age, y=friend_count, data=pf)
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_point() +
  xlim(13,90)
summary(pf$age)
```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_jitter(alpha=1/20) +
  xlim(13,90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_point(alpha=1/20) + xlim(13,90) + coord_trans(y='sqrt')
ggplot(aes(x=age, y=friend_count), data=pf) +
  geom_point(alpha=1/20, position = position_jitter(h=0)) + xlim(13,90) + coord_trans(y='sqrt')
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated), data=pf) +
  geom_jitter(alpha=1/10, position = position_jitter(h=0)) +
  coord_trans(y='sqrt')
  
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
library(dplyr)
age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups,
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)

#or
pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age)

head(pf.fc_by_age) 

```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +geom_line()
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x=age, y=friendships_initiated), data=pf) +
  coord_cartesian(xlim=c(13,70), ylim=c(0,1000)) +
  geom_point(alpha=0.05, position = position_jitter(h=0), color = 'orange') +
  geom_line(stat='summary', fun.y=mean) +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.1), linetype=2, color='blue') +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.5), color='blue') +
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.9), linetype=2, color='blue')

```

#### What are some of your observations of the plot?
Response:
Q3 always under 1000 friends

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor(pf$age, pf$friend_count)
?cor
with(pf, cor.test(age, friend_count, method='pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count, method='pearson'))
with(subset(pf, age <= 70), cor.test(age, friend_count, method='kendall'))
with(subset(pf, age <= 70), cor.test(age, friend_count, method='spearman'))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) +
  geom_point(alpha=1/20, position = position_jitter(h=0)) +
  scale_x_sqrt(limits=c(0,10000)) +
  scale_y_sqrt(limits=c(0,10000))
?scale_x_sqrt
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x=www_likes_received, y=likes_received), data=pf) +
  geom_point(alpha=1/10) +
  xlim(0, quantile(pf$www_likes_received, 0.95)) + 
  ylim(0, quantile(pf$likes_received, 0.95)) + 
  geom_smooth(method='lm', color='red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(likes_received, www_likes_received, method='pearson'))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
data("Mitchell")
str(Mitchell)
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(Month, Temp), data=Mitchell) +
  geom_point() +
  scale_x_continuous(breaks=seq(0,203,12))
with(Mitchell, cor.test(Month, Temp, method='pearson'))
ggplot(aes(x=(Month%%12),y=Temp), data=Mitchell)+
  geom_point()

install.packages('energy')
library(energy)
?dcor.ttest

x <- seq(0, 4*pi, pi/20)
y <- cos(x)
qplot(x = x, y = y)
dcor.ttest(x, y)
cor.test(x,y)

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x=age, y=friend_count_mean), data=pf.fc_by_age) +geom_line()
head(pf.fc_by_age, 10)
pf.fc_by_age[17:19,]

names(pf)
pf$age_with_months <- pf$age + (12 - pf$dob_month)/12

```

***

### Age with Months Means

```{r Age with Months Means}
library(dplyr)

age_with_months_groups <- group_by(pf, age_with_months)
pf.fc_by_age_months <- summarise(age_with_months_groups, 
                             friend_count_mean = mean(friend_count), 
                             friend_count_median = median(friend_count), 
                             n = n())
pf.fc_age_month <- arrange(pf.fc_by_age_month, age_with_months)
head(pf.fc_age_month)

#or
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age_with_months)

head(pf.fc_by_age_months)
```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
p1 <- ggplot(aes(age, friend_count_mean), data =subset(pf.fc_by_age, age<71)) +
  geom_line()

p2 <- ggplot(aes(age_with_months, friend_count_mean), data=subset(pf.fc_by_age_months, age_with_months <71)) +
  geom_line()

library(gridExtra)
grid.arrange(p2, p1, ncol=1)
```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(aes(age, friend_count_mean), data =subset(pf.fc_by_age, age<71)) +
  geom_line() + 
  geom_smooth()

p2 <- ggplot(aes(age_with_months, friend_count_mean), data=subset(pf.fc_by_age_months, age_with_months <71)) +
  geom_line() +
  geom_smooth()

grid.arrange(p2, p1, ncol=1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:
new correlation measures: spearman, Kendall...

ttest of multivariate independence possible in r with dcor.ttest()

how to show multiple plots on one figure with gridExtra package and grid.arrange()
how to overlay summaries with raw data with geom_xx(stat='summary', fun.y=... fun.args=..)
how to avoid error message geom_jitter with geom_jitter(position=position_jitter(h=0))
learn about dplyr package and how to do conditional summaries with group_by and chain statements with %>%
use of transparency (alpha) and jitter to reduce overplotting


***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

